Gompertz growth as infinitely correlated non-local growth

### 2024-04-18 arxiv2/new_draft.tex
first modification

\begin{abstract}
Logistic and Gompertz growth has traditionally been connected by augmenting the logistic model with an extra parameter to obtain what is known as $\theta$-logistic growth or the Richards model for growth. In spite of this bridge, the biological foundation of Gompertz growth is only vaguely understood. 
To add to this biological foundation, we show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth where the latter involves a network of entities. 
The microscopic representation elucidates the source of Gompertz growth as the potential abundance and not the realized abundance, and an example of such a network is presented. 
For finite higher order terms we show that the growth equation can be represented by Gauss' Hypergeometric function, and only for infinitely many terms does the abundance lend itself conveniently to a log-transformation wherein the log-abundance growth rate decreases linearly, the trademark of Gompertz growth. We thus equate Gompertz growth with correlated microscopic entities that exhibit long-range or non-local interaction. 
This conclusion adds to the interpretation of biological systems that undergo Gompertz growth where long-range interaction could be interpreted as the source of growth being external to the individual entities, like a field that stimulates growth simultaneously across all entities. 
This field could be interpreted as an environmental stressor like air-pollution, age, radiation, etc, or an environmental stimulant like an electromagnetic field. 
Our new discovery allows for a unification of well-known growth models as follows: time independent and uncorrelated growth yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and time independent and infinitely correlated growth yields the Gompertz function.

@@

\subsection{Main Results and Roadmap}
In this paper, our primary aim is to investigate the source of growth in self-organized processes. Second, we will elaborate on the idea that the logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth. We will devise a slightly different parametrization than the Richards model, the generalized Gompertz-Logistic equation which is commonly used to compare the two models (e.g. by \citet{petroni2020logistic}, \citet{tjorve2017use}, and \citet{wang2012richards}). In particular, instead of considering a decimal number as the unifying exponent we will appeal to integer exponents with a number of terms that tend towards infinity as we approach the Gompertz model. As we shall see, this is rooted in the microscopic interpretation of the growth system, where each subgroup of $N$ elementary entities interacts with all other equally sized subgroups each composed of mutually exclusive elementary entities. If these subgroups are present at all orders of cardinality, the Gompertz model emerges, while only pairwise interactions yield the logistic model. As seen in the aforementioned references, this is in line both with experimental and theoretical results. In presenting this argument, we present a novel bridge between the logistic and the Gompertz model which offers both a mathematical and biological foundation to the source of growth. Lastly, we will present a unifying framework in the growth context for the common functions Gaussian, Exponential, Logistic, Gompertz and Poisson.

@@

First consider a system that grows according to logistic growth. If we let the variable $X(t)$ denote the abundance of a quantity in a normalized system where its maximum size is 1, the equation that determines its size as a function of time is
\begin{equation}
  \frac{1}{X}\frac{d}{dt} X(t) = \beta (1 - X(t)),
\end{equation}
where $\beta>0$ is a growth parameter. Here, the right hand side is commonly interpreted as representing ``resources available for growth'' which decrease linearly with respect to the relative abundance's growth rate on the left hand side. One way to generalize logistic growth is to introduce a parameter that exponentiates the resource availability in the following way,
\begin{equation}
\label{eq:Rich}
  \frac{1}{X}\frac{d}{dt} X(t) = \beta (1 - X^{\theta}(t)),
\end{equation}
also called $\theta$-logistic growth or Richards model \citep{richards1959flexible}. Petroni et al. (2020) argued that this exponentiation can be interpreted as non-linear interaction effects where there is a level of cooperation in the microscopic domain. They showed that in the limit where the system becomes \textit{maximally cooperative} or \textit{coherent}, $\theta\rightarrow 0$ and the $\theta$-logistic equation reduces in this limit to the Gompertz growth model,
 \begin{equation}
 \label{eq:Gomp}
  \frac{1}{X}\frac{d}{dt} X(t) = - \beta \ln X(t).
\end{equation}
In this limit, the relative growth goes to infinity as $\lim_{t\to 0}$, which could be seen as a verification of the maximally coherent characteristic of the system. One is reminded of a system that is perturbed and left to reach a new equilibrium. Petroni's definition of maximal coherence in the context of the Gompertz equation was shown to rigorously correspond to the definition of coherence used in quantum mechanical systems \cite{molski2003coherent} attributed to Glauber (1963) \cite{glauber1963coherent}.

@@

 The emergence of Gompertz growth from higher order logistic growth can be seen as another way to encode non-linear effects that were seen as coherence in accordance with Petroni and Molski. In practice, the augmented logistic model has a faster growth rate decay than the familiar $\theta$-logistic growth (Figure \ref{fig:hypergeometric_plain}). If we interpret deceleration of relative growth as degree of coherence, the augmented logistic growth would be more coherent than the $\theta$-logistic model for a given starting value.

@@

Now consider the growth equation with only pairwise interactions, or \textit{first order} interactions, governed by an adjacency matrix ${a_{ij}}$ as a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise, viz.
\begin{equation}
\label{eq:modSIR}
\frac{d x_i}{dt} = x_i\sum_{j \in \mathcal{N}(i)}{a_{ij}}s_{j} \quad \forall i,
\end{equation}
or
\begin{equation}
\label{eq:modSIR2}
\frac{d}{dt}\ln x_i = \sum_{j \in \mathcal{N}(i)}{a_{ij}}s_{j} \quad \forall i,
\end{equation}
where $\mathcal{N}(i)$ is the neighborhood of nodes connected to node $i$. For example, if the adjacency matrix has only $1$s and no $0$s, this means that every node is connected to every other node. The diagonal of the adjacency matrix is zero as a node is not connected to itself by definition.

@@

Now as we let N grow, we impose a simple rule: \textit{Each entity can only receive growth potential from a single entity and create growth potential in a single possibly different entity}. This condition translates into an adjacency matrix with each column containing a single non-zero entry. 
Using this rule and rearranging and summing across all entities we obtain the following relationship,
\begin{equation}
\label{eq:modSIR2}
\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ] = \sum_j^N s_{j}.
\end{equation}

@@

\subsection{The invariance of logistic growth under causality reversal}
\label{appendix:swappinglogistic}
Consider a causality reversal of logistic growth such that the realized growth drives potential growth. Then Equation \ref{eq:modSIR} must be modified by swapping $x$ and $s$ on the RHS as follows,
\begin{equation}
\label{eq:revSIR}
\frac{d x_i}{dt} = s_i\sum_{j \in \mathcal{N}(i)}{a_{ij}}x_{j} \quad \forall i.
\end{equation}
Using the identity,
\begin{equation}
\label{eq:modSIR2}
\sum_i\frac{1}{s_i}\frac{dx_i}{dt} = -\sum_i\frac{1}{s_i}\frac{ds_i}{dt} = -\frac{d}{dt} \ln \left[ \prod_i^N s_i \right ],
\end{equation}
and using the equivalent connectivity rule as outlined in Section \ref{sec:micro}, we have,
\begin{equation}
\label{eq:rev_id}
-\frac{d}{dt} \ln \left[ \prod_i^N s_i \right ] = \sum_j^N x_j.
\end{equation}

>>>>

\begin{abstract}
Logistic and Gompertz growth has traditionally been connected by augmenting the logistic model with an extra parameter to obtain what is known as $\theta$-logistic growth or the Richards model for growth. In spite of this bridge, the biological foundation of Gompertz growth is only vaguely understood. 
To add to this biological foundation, we show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth where the latter involves a network of entities. 
The microscopic representation elucidates the source of Gompertz growth as the potential abundance and not the realized abundance, and an example of such a network is presented. 
For finite higher order terms we show that the growth equation can be represented by Gauss' Hypergeometric function, and only for infinitely many terms does the abundance lend itself conveniently to a log-transformation wherein the log-abundance growth rate decreases linearly, the trademark of Gompertz growth. We thus equate Gompertz growth with correlated microscopic entities that exhibit long-range or non-local interaction. 
This conclusion adds to the interpretation of biological systems that undergo Gompertz growth where long-range interaction could be interpreted as the source of growth being external to the individual entities, like a field that stimulates growth simultaneously across all entities. 
This field could be interpreted as an environmental stressor like air-pollution, radiation, etc, or an environmental stimulant like an electromagnetic field. 
Our new discovery allows for a unification of well-known growth models as follows: time independent and uncorrelated growth yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and time independent and infinitely correlated growth yields the Gompertz function.

@@

\subsection{Main Results and Roadmap}
In this paper, our primary aim is to investigate the source of growth in self-organized processes. In particular, we posit that the source of growth can be elucidated by considering interactions in the microscopic domain which can be aggregated to recover the traditional macroscopic growth equations. Second, we will elaborate on the idea that the logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth. We will devise a slightly different parametrization than the Richards model, the generalized Gompertz-Logistic equation which is commonly used to compare the two models (e.g. by \citet{petroni2020logistic}, \citet{tjorve2017use}, and \citet{wang2012richards}). In particular, instead of considering a decimal number as the unifying exponent we will appeal to integer exponents with a number of terms that tend towards infinity as we approach the Gompertz model. As we shall see, this is rooted in the microscopic interpretation of the growth system, where each subgroup of $N$ elementary entities interacts with all other equally sized subgroups each composed of mutually exclusive elementary entities. If these subgroups are present at all orders of cardinality, the Gompertz model emerges, while only pairwise interactions yield the logistic model. As seen in the aforementioned references, this is in line both with experimental and theoretical results. In presenting this argument, we present a novel bridge between the logistic and the Gompertz model which offers both a mathematical and biological foundation to the source of growth. Lastly, we will present a unifying framework in the growth context for the common functions Gaussian, Exponential, Logistic, Gompertz and Poisson.

@@

First consider a system that grows according to logistic growth. If we let the variable $X(t)$ denote the abundance of a quantity in a normalized system where its maximum size is 1, the equation that determines its size as a function of time is
\begin{equation}
  \frac{1}{X}\frac{d}{dt} X(t) = \beta (1 - X(t)),
\end{equation}
where $\beta>0$ is a growth parameter. Here, the right hand side is commonly interpreted as representing ``resources available for growth'' which decrease linearly with respect to the relative abundance's growth rate on the left hand side. One way to generalize logistic growth is to introduce a parameter that exponentiates the resource availability in the following way,
\begin{equation}
\label{eq:Rich}
  \frac{1}{X}\frac{d}{dt} X(t) = \frac{\beta}{\theta} (1 - X^{\theta}(t)),
\end{equation}
also called $\theta$-logistic growth or Richards model \citep{richards1959flexible}. Petroni et al. (2020) argued that this exponentiation can be interpreted as non-linear interaction effects where there is a level of cooperation in the microscopic domain. They showed that in the limit where the system becomes \textit{maximally cooperative} or \textit{coherent}, $\theta\rightarrow 0$ and the $\theta$-logistic equation reduces in this limit to the Gompertz growth model,
 \begin{equation}
 \label{eq:Gomp}
  \frac{1}{X}\frac{d}{dt} X(t) = - \beta \ln X(t).
\end{equation}
In this limit, the relative growth goes to infinity as $\lim_{t\to 0}$, which could be seen as a verification of the maximally coherent characteristic of the system. One is reminded of a system that is perturbed and left to reach a new equilibrium. Petroni's definition of maximal coherence in the context of the Gompertz equation was shown to rigorously correspond to the definition of coherence used in quantum mechanical systems \cite{molski2003coherent} attributed to Glauber (1963) \cite{glauber1963coherent}.

@@

 The emergence of Gompertz growth from higher order logistic growth can be seen as another way to encode non-linear effects that were seen as coherence in accordance with Petroni and Molski. In practice, the augmented logistic model has a slower growth rate decay than the familiar $\theta$-logistic growth (Figure \ref{fig:hypergeometric_plain}). If we interpret deceleration of relative growth as degree of coherence, the augmented logistic growth would be more coherent than the $\theta$-logistic model for a given starting value.

@@

Now consider the growth equation with only pairwise interactions, or \textit{first order} interactions, governed by an adjacency matrix ${a_{ij}}$ as a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise, viz.
\begin{equation}
\label{eq:modSIR}
\frac{d x_i}{dt} = x_i\sum_{j \in \mathcal{N}(i)}{a_{ij}}s_{j} \quad \forall i,
\end{equation}
or
\begin{equation}
\label{eq:modSIR2}
\frac{d}{dt}\ln x_i = \sum_{j \in \mathcal{N}(i)}{a_{ij}}s_{j} \quad \forall i,
\end{equation}
where $\mathcal{N}(i)$ is the neighborhood of nodes connected to node $i$, excluding node $i$ itself. For example, if the adjacency matrix has only $1$s and no $0$s, this means that every node is connected to every other node. The diagonal of the adjacency matrix is zero as a node is not connected to itself by definition.

@@

Now as we let N grow, we impose a simple rule: \textit{Each entity can only receive growth potential from a single entity and create growth potential in a single possibly different entity}. This condition translates into an adjacency matrix with each column containing a single non-zero entry. 
Using this rule and rearranging and summing across all entities we obtain the following relationship,
\begin{equation}
\label{eq:modSIR22}
\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ] = \sum_j^N s_{j}.
\end{equation}

@@

\subsection{The invariance of logistic growth under causality reversal}
\label{appendix:swappinglogistic}
Consider a causality reversal of logistic growth such that the realized growth drives potential growth. Then Equation \ref{eq:modSIR} must be modified by swapping $x$ and $s$ on the RHS as follows,
\begin{equation}
\label{eq:revSIR}
\frac{d x_i}{dt} = s_i\sum_{j \in \mathcal{N}(i)}{a_{ij}}x_{j} \quad \forall i.
\end{equation}
Using the identity,
\begin{equation}
\label{eq:modSIR2app}
\sum_i\frac{1}{s_i}\frac{dx_i}{dt} = -\sum_i\frac{1}{s_i}\frac{ds_i}{dt} = -\frac{d}{dt} \ln \left[ \prod_i^N s_i \right ],
\end{equation}
and using the equivalent connectivity rule as outlined in Section \ref{sec:micro}, we have,
\begin{equation}
\label{eq:rev_id}
-\frac{d}{dt} \ln \left[ \prod_i^N s_i \right ] = \sum_j^N x_j.
\end{equation}
